Level-Set method on \texttt{p4est}

### 2015-02-04 paper/tex/parallel_algorithms.tex
final pass (mohammad+frederic+arthur) done. One last thing remains, scaling with large number of processes and for a heavier load

Once the gird is adapted to the interface, it must be partitioned to ensure load balancing across processes. This is achieved by constructing a Z-curve that traverses the leaves of all trees in order of tree index (c.f. figure \ref{fig:p4est_zcurve}). A Z-curve is a Space Filling Curve (SFC) with the important property that cells with close Z-indices are also geometrically close in the physical domain. This is beneficial since it leads to both reduction in MPI communications and improvements of the cache performance of several algorithms such as interpolation and finite difference calculations. For more details on parallel partitioning in \texttt{p4est} one may refer to section 3.3 of \cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo}.
\begin{figure}[hbtp]
\begin{center}
\includegraphics[width = \columnwidth]{figures/p4est_zcurve.pdf}
\caption{Left: a ``forest'' made up of two trees $k_0$ and $k_1$. Parallel partitioning is achieved by first constructing a Z-curve starting at cell ``S'' and ending at cell ``E''. Next, the one dimensional curve is split up among processes either uniformly or by assigning different weights to cells. Here processes are represented via different colors. Note how using the Z-curve naturally leads to clustering of most cells in each domain. Right: schematic of a tree data structure representing the forest and its partitioning.}
\label{fig:p4est_zcurve}
\end{center}
\end{figure}
Aside from grid manipulation and partitioning, we use two additional features of \texttt{p4est}, namely the generation of ghost layer cells and the creation of a globally unique node indexing. These algorithms are detailed in sections 3.5 and 3.6 of \cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo}. An important feature of our algorithms is that they are designed for non-graded trees. This is important because we can entirely skip tree balancing, which was shown to be one of the most time consuming parts of grid adaptation in \texttt{p4est} \cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo}.

Finally, in \texttt{p4est} trees are linearized, i.e. only the leaves are explicitly stored. However, explicit knowledge of the hierarchal structure of the tree is greatly beneficial in several algorithms, e.g. in search operations needed for the interpolation algorithm. Thus, we introduce a simple reconstruction algorithm that recreates a local representation of the entire ``forest'' that is only adapted to local cells and, potentially, the ghost layer. This approach is similar to the ideas introduced in \cite{Bangerth;Burstedde;Heister;etal:11:Algorithms-and-data-} and our tests show that in a typical application they amount to less that $1\%$ of the entire runtime. Algorithm \ref{alg:reconstruction} illustrates how this reconstruction is performed. Given a forest and a layer of ghost cells from \texttt{p4est}, the algorithm generates a local representation of the forest by recursively refining from the root until reaching the same level and location of all leaves in the local forest and the ghost layer. Note that algorithm \ref{alg:reconstruction} does not involve any communication and is load balanced provided that the initial forest is balanced. Figure \ref{fig:reconstruction} illustrates an example where Algorithm \ref{alg:reconstruction} is applied. Note how each process has independently generated a local representation of the forest that is refined to match the same leaves as in the global forest and ghost layer.
\begin{figure}[htbp]
\begin{center}
\includegraphics[width = \textwidth]{figures/reconstruct.pdf}
\end{center}
\caption{Left: a forest refined close to an interface and partitioned among four processes, as indicated by colors. Right: each process independently recreates a local forest that is refined to match the local grid and is as coarse as possible elsewhere. Note that empty cells are fictitious, i.e. they are only required to generate the hierarchal structure and are not matched by any corresponding cell in the global forest.}
\label{fig:reconstruction}
\end{figure}
\begin{algorithm}[htbp]
\caption{$H \gets \texttt{Reconstruct (}G\texttt{)}$} %\label{euclid}
\begin{algorithmic}[1]
\State $H \gets G.\texttt{macromesh()}$
\For {$tr : G.\texttt{local\_trees()}$} \Comment{traverse local cells}
	\For {$q : tr.\texttt{cells()}$}
		\State $H.\texttt{update\_tree(} tr, q \texttt{)}$
	\EndFor
\EndFor

\For {$q : G.\texttt{ghost\_cells()}$} \Comment{traverse ghost cells}
	\State $H.\texttt{update\_tree(} q.\texttt{tree()}, q \texttt{)}$
\EndFor

\State \Return $H$
\\
\Function{$H.\texttt{update\_tree}$}{$tr, q$}   \Comment{recursive reconstruction}
	\State $q_l \gets H.\texttt{root(}tr\texttt{)}$
	\While{$q_l.\texttt{level()} \not= q.\texttt{level()} $}
		\If {$q_l.$\texttt{is\_leaf()}} $q_l$\texttt{.split()}
		\EndIf
		\State $h \gets q_l.\texttt{length()} / 2$ \Comment{select the next child based on direction}
		\State $i \gets q.x \ge q_l.x + h$
		\State $j \gets q.y \ge q_l.y + h$
		\State $k \gets q.z \ge q_l.z + h$

		\State $q_l \gets q_l.\texttt{child(} i, j, k \texttt{)}$
	\EndWhile	
\EndFunction
\end{algorithmic}
\label{alg:reconstruction}
\end{algorithm}

\subsection{Interpolation and semi-Lagrangian methods}
As indicated earlier, we use the semi-Lagrangian method to solve equation \eqref{eq:ls} when the velocity field is externally generated, i.e. when it does not depend explicitly on the level-set function itself. Let us rewrite equation \eqref{eq:ls} along the characteristic curve $\mathbf{X}(t)$ as:
\be
\left\{
\begin{array}{rcl}
\dfrac{\text{d} \mathbf{X}}{\text{d} t} &=& \underline{\mbf{u}}, \\ [3ex]
\dfrac{\text{d} \phi(\mathbf{X}(t), t)}{\text{d} t} &=& 0.
\end{array}
\right.
\label{eq:sl}
\ee
The semi-Lagrangian method integrates equations \eqref{eq:sl} backward in time, i.e. starting from the grid $G^{n+1}$ \footnote{$G^{n+1}$ itself is computed in an iterative fashion as explained later on.}, we simply write $\phi^{n+1}(\mathbf{X}^{n+1}) = \phi(\mathbf{X}(t^{n+1}), t^{n+1}) = \phi(\mathbf{X}(t^n), t^n) = \phi^n(\mathbf{X}_d)$. Here, the characteristic curves are chosen such that $\mathbf{X}(t^{n+1})$ are simply the coordinates of grids points of $G^{n+1}$, and $\mathbf{X}_d$ are the departure points, which are computed using the second-order midpoint method \cite{Min;Gibou:07:A-second-order-accur}:
\bea
\mathbf{X}^\star &=& \mathbf{X}^{n+1} - \frac{\Delta t}{2} \underline{\mbf{u}}^{n}(\mathbf{X}^n),	   \label{eq:xstar}      \\
\mathbf{X}_d     &=& \mathbf{X}^{n+1} - \Delta t \underline{\mbf{u}}^{n+\frac{1}{2}}(\mathbf{X}^\star), \label{eq:xdeparture}
\eea
where $\underline{\mbf{u}}^{n+\frac{1}{2}}$ is obtained via extrapolation from previous times, i.e.:
\be
\underline{\mbf{u}}^{n+\frac{1}{2}} = \frac{3}{2} \underline{\mbf{u}}^n - \frac{1}{2}\underline{\mbf{u}}^{n-1}. \label{eq:vn_p_half}
\ee
Note that all values at the intermediate point, $\mathbf{X}^\star$, and departure point, $\mathbf{X}_d$, must be calculated via interpolation from the previous grids $G^{n}$ and $G^{n-1}$. Here, we use the stabilized second-order interpolation for $\phi(\mathbf{X}_d)$ and the multi-linear interpolation for $\underline{\mbf{u}}^{n+\frac{1}{2}}(\mathbf{X}^\star)$ \cite{Min;Gibou:07:A-second-order-accur}. Although parallelization of the interpolation process on a shared-memory machine is trivial, the same cannot be said for distributed-memory machines. In fact, the parallel interpolation procedure given in Algorithm \ref{alg:interpolation} is probably the most important contribution of this article since the procedure, which is trivial on uniform grids, is challenging in the case of trees because it is not straightforward to identify which processes owns the departure points. Indeed, complications arise because not all departure points will reside in the domain owned by the current process. Moreover, due to the irregular shapes of the partitions, one cannot even ensure they are entirely owned by neighboring processes. At best we can only expect that their locations are bounded by a halo of width $w \le \text{CFL} \: \Delta x_{\min}$ around the local partition, where $x_{\min}$ is the size of the smallest cell in the forest. Naturally, if one enforces $\text{CFL} \le 1$, one can ensure that the halo is bounded by the ghost layer, which significantly simplifies the communication problem. This assumption, however, defeats the purpose of using a semi-Lagrangian approach, whose purpose is to enable large CFL values.

One remedy to this problem, proposed in \cite{Thomas;Cote:95:Massively-parallel-s} for uniform grids, is to increase the size of ghost layer to $\lceil \text{CFL} \rceil$. For large values of the $\text{CFL}$ number, however, this approach can substantially increase the communication volume. Moreover, this simple approach does not work in the process of generating $G^{n+1}$ due to repartitioning. An alternative approach would be to handle local and remote interpolations separately. Our remote interpolation algorithm is composed of three separate phases. In the first phase, which we call buffering, every process searches for all departure points inside the local trees. If the point is owned by a local cell, it is added to a local buffer, otherwise we find the process which owns the point and add the point to a separate buffer belonging to the found rank. Note that searching the point in the local tree is performed recursively using the hierarchal reconstruction (c.f.\ algorithm \ref{alg:reconstruction}). Moreover, the owner's rank is found by computing the Z-index of the point and then using a binary search on the Z-curve. This is already implemented in \texttt{p4est} and explained in details in section 2.5 of \cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo}.

@@

To solve the communication problem, we first compute the adjacency matrix of the communication pattern, i.e. we construct the matrix $A_{P \times P}$, where $P$ is the number of processes, such that
\ben
a_{ij} = 
\left\{
\begin{array}{lr}
1 \hspace{5 mm} \text{if process `\textit{i}' sends a message to process `\textit{j}',} \\
0 \hspace{5 mm} \text{otherwise.}
\end{array}
\right.
\een
Note that this matrix is also distributed among processes, i.e. each row is owned by a separate rank. Next, we compute
\ben
S_i = \sum_j a_{ij} \hspace{5 mm} \text {and} \hspace{5 mm} R_i = \sum_j a_{ji}
\een
where $S_i$ and $R_i$ denote the number of sent and received messages, respectively. While $S_i$ can be computed trivially, a reduction operation is required to compute $R_i$. For instance, this can be achieved using a single \texttt{MPI\_Reduce\_scatter} function call. The last phase of the interpolation procedure involves overlapping the computation of interpolated values for local points with the communication of data between processes. This is done by alternating between local calculations and probing for incoming messages from other processes. The interpolation is finished once the values for all local the points have been calculated and all the remote requests have been processed (see Algorithm \ref{alg:interpolation}).
\begin{algorithm}[htbp]
\caption{$values \gets \texttt{Interpolate (}H, \mathbf{X}\texttt{)}$} %\label{euclid}
\begin{algorithmic}[1]
\State $col \gets 0, buff \gets null$ \Comment{Phase I -- buffering}
\For {$\mathbf{p} : \mathbf{X}$} 
	\State $[r, cell] \gets H.\texttt{search(}\mathbf{p}\texttt{)}$ \Comment{search for the owner's rank and cell}
	\If {$r = mpirank$}
		\State $buff[r].\texttt{push\_back(}\mathbf{p}, cell\texttt{)}$
	\Else
		\State $buff[r].\texttt{push\_back(}\mathbf{p}\texttt{)}$
		\State $col[r] \gets 1$
	\EndIf
\EndFor
\For {$r: mpisize$} \Comment{Phase II -- initiate communication and compute number of messages}
	\If {$col[r]$}
		\State $\texttt{MP\_Isend(}r, buff[r]\texttt{)}$
	\EndIf
\EndFor
\State  $S \gets \texttt{sum(}col\texttt{)}$ 
\State  $R \gets \texttt{MPI\_Reduce\_scatter(}col,\texttt{MPI\_SUM)}$
\State $done \gets false, it \gets buff[mpirank].\texttt{begin()}$
\While {$!done$} \Comment{Phase III -- main loop}
	\If {$it \not= buff[mpirank].\texttt{end()}$}
		\State $values \gets \texttt{process\_local(}it\texttt{)}$ \Comment{process local interpolations}
		\State $\texttt{++}it$
	\EndIf
	\If {$R > 0$} \Comment{process queries sent from remote processes}
		\State $[msg, st] \gets \texttt{MPI\_Iprobe()}$
		\If {$msg$}
			% \State $values \gets \texttt{MPI\_Recv(}st.\texttt{MPI\_SOURCE)}$ \Comment{}
			\State $val\_buff \gets \texttt{process\_queries(}st\texttt{)}$ \Comment{receive, search, and interpolate values}
			\State $\texttt{MPI\_Isend(}st.\texttt{MPI\_SOURCE},val\_buff\texttt{)}$ \Comment{send back interpolated values}
			\State $R\texttt{--}$
		\EndIf
	\EndIf
	\If {$S > 0$} \Comment{process replies sent to our queries}
		\State $[msg, st] \gets \texttt{MPI\_Iprobe()}$
		\If {$msg$}
			\State $values \gets \texttt{process\_replies(}st.\texttt{MPI\_SOURCE)}$ \Comment{receive remotely interpolated values}
			\State $S\texttt{--}$
		\EndIf
	\EndIf
	\State $done \gets S = 0 \: \And \: R = 0 \: \And \: it = buff[mpirank].\texttt{end()}$
\EndWhile
\State \Return $values$
\end{algorithmic}
\label{alg:interpolation}
\end{algorithm}

Using the interpolation Algorithm \ref{alg:interpolation}, we close this section by presenting the final semi-Lagrangian Algorithm \ref{alg:semi-lagrangian}. The basic idea is to start from an initial guess $G^{n+1}_0$ for the grid and modify it using the refinement \eqref{eq:refine} and coarsening \eqref{eq:coarsen} criteria until convergence is obtained. Various options are available for $G^{n+1}_0$. For instance it is possible to start from the macromesh and only perform refinement steps until convergence. This choice, however, is not suitable since the first few iterations do not contain many cells and there is little work for parallelism. Here we simply take the previous grid as the starting point, i.e. $G^{n+1}_0 = G^n$. Note that this iterative process is essentially unavoidable since the grid is based on the values of the level-set function at $t^{n+1}$, which itself is unknown and is to be defined on $G^{n+1}$. Nonetheless the process converges to the final grid in at most $l_{\max}-l_{\min}$ steps where $l_{\min}$ and $l_{\max}$ denote the maximum and minimum depth of all trees in the forest, receptively.
\begin{algorithm}[htbp]
\caption{$[G^{n+1}, \phi^{n+1}] \gets \texttt{SemiLagrangian (}G^n, \phi^n, \mathbf{V}^n, \mathbf{V}^{n-1}, \text{CFL}\texttt{)}$} %\label{euclid}
\begin{algorithmic}[1]
\State $\Delta t_l \gets \text{CFL} \times G^n.\texttt{hmin()} / \max \{\mathbf{V}^n \} $
\State $\Delta t   \gets \texttt{MPI\_Allreduce(}\Delta t_l,\texttt{MPI\_MIN)}$
\State $H^n \gets \texttt{Reconstruct(} G^n\texttt{)}$ 
\State $G^{n+1}_0 \gets G^n$
\While {$true$}
	\State $\mathbf{X}_d  \gets \texttt{ComputeDeparturePoints(}G^{n+1}_0, \mathbf{V}^n, \mathbf{V}^{n-1}, \Delta t\texttt{)}$ \Comment{using equations \ref{eq:xstar} -- \ref{eq:vn_p_half}}
	\State $\phi^{n+1} \gets \texttt{Interpolate(}H^n, \textbf{X}_d\texttt{)}$
	\State $G^{n+1} \gets G^{n+1}_0.\texttt{refine\_and\_coarsen(}\phi^{n+1}\texttt{)}$ \Comment{using equations \ref{eq:refine} and \ref{eq:coarsen} as criteria}
	\If {$G^{n+1} \not= G^{n+1}_0$}
		\State $G^{n+1}.\texttt{partition()}$
		\State $G^{n+1}_0 \gets G^{n+1}$
	\Else
		\State \textbf{break}
	\EndIf
\EndWhile

>>>>

Once the gird is adapted to the interface, it must be partitioned to ensure load balancing across processes. This is achieved by constructing a Z-curve that traverses the leaves of all trees in order of tree index (c.f. figure \ref{fig:p4est_zcurve}). A Z-curve is a Space Filling Curve (SFC) with the important property that cells with close Z-indices are also geometrically close in the physical domain. This is beneficial since it leads to both reduction in MPI communications and improvements of the cache performance of several algorithms such as interpolation and finite difference calculations. For more details on parallel partitioning in \texttt{p4est} one may refer to section 3.3 of \cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo}.
\begin{figure}[hbtp]
\begin{center}
\includegraphics[width = \columnwidth]{figures/p4est_zcurve.pdf}
\caption{Left: a ``forest'' made up of two trees $T_0$ and $T_1$. Parallel partitioning is achieved by first constructing a Z-curve starting at cell ``S'' and ending at cell ``E''. Next, the one dimensional curve is split up among processes either uniformly or by assigning different weights to cells. Here processes are represented via different colors. Note how using the Z-curve naturally leads to clustering of most cells in each domain. Right: schematic of a tree data structure representing the forest and its partitioning.}
\label{fig:p4est_zcurve}
\end{center}
\end{figure}
Aside from grid manipulation and partitioning, we use two additional features of \texttt{p4est}, namely the generation of ghost layer cells and the creation of a globally unique node indexing. These algorithms are detailed in sections 3.5 and 3.6 of \cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo}. An important feature of our algorithms is that they are designed for non-graded trees. This is important because we can entirely skip tree balancing, which was shown to be one of the most time consuming parts of grid adaptation in \texttt{p4est} \cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo}.

Finally, in \texttt{p4est} trees are linearized, i.e. only the leaves are explicitly stored. However, explicit knowledge of the hierarchal structure of the tree is greatly beneficial in several algorithms, e.g. in search operations needed for the interpolation algorithm. Thus, we introduce a simple reconstruction algorithm that recreates a local representation of the entire ``forest'' that is only adapted to local cells and, potentially, the ghost layer. This approach is similar to the ideas introduced in \cite{Bangerth;Burstedde;Heister;etal:11:Algorithms-and-data-} and our tests show that in a typical application they amount to less that $1\%$ of the entire runtime. Algorithm \ref{alg:reconstruction} illustrates how this reconstruction is performed. Given a forest and a layer of ghost cells from \texttt{p4est}, the algorithm generates a local representation of the forest by recursively refining from the root until reaching the same level and location of all leaves in the local forest and the ghost layer. Note that algorithm \ref{alg:reconstruction} does not involve any communication and is load balanced provided that the initial forest is balanced. Figure \ref{fig:reconstruction} illustrates an example where Algorithm \ref{alg:reconstruction} is applied. Note how each process has independently generated a local representation of the forest that is refined to match the same leaves as in the global forest and ghost layer.
\begin{figure}[htbp]
\begin{center}
\includegraphics[width = \textwidth]{figures/reconstruct.pdf}
\end{center}
\caption{Left: a forest refined close to an interface and partitioned among four processes, as indicated by colors. Right: each process independently recreates a local forest that is refined to match the local grid and is as coarse as possible elsewhere. Note that empty cells are fictitious, i.e. they are only required to generate the hierarchal structure and are not matched by any corresponding cell in the global forest.}
\label{fig:reconstruction}
\end{figure}
\begin{algorithm}[htbp]
\caption{$H \gets \texttt{Reconstruct (}G\texttt{)}$: construction of the local tree hierarchy $H$ from the grid $G$}
\begin{algorithmic}[1]
\State $H \gets G.\texttt{macromesh()}$
\For {$tr : G.\texttt{local\_trees()}$} \Comment{build hierarchy for local cells}
	\For {$c : tr.\texttt{cells()}$}
		\State $H.\texttt{update\_tree(} tr, c \texttt{)}$
	\EndFor
\EndFor

\For {$c : G.\texttt{ghost\_cells()}$} \Comment{build hierarchy for ghost cells}
	\State $H.\texttt{update\_tree(} c.\texttt{tree()}, c \texttt{)}$
\EndFor

\State \Return $H$
\\
\Function{$H.\texttt{update\_tree}$}{$tr, c$}   \Comment{recursive tree reconstruction}
	\State $c_l \gets H.\texttt{root(}tr\texttt{)}$
	\While{$c_l.\texttt{level()} \not= c.\texttt{level()} $}
		\If {$c_l.$\texttt{is\_leaf()}} $c_l$\texttt{.split()}
		\EndIf
		\State $h \gets c_l.\texttt{length()} / 2$ \Comment{select the next child based on direction}
		\State $i \gets c.x \ge c_l.x + h$
		\State $j \gets c.y \ge c_l.y + h$
		\State $k \gets c.z \ge c_l.z + h$

		\State $c_l \gets c_l.\texttt{child(} i, j, k \texttt{)}$
	\EndWhile	
\EndFunction
\end{algorithmic}
\label{alg:reconstruction}
\end{algorithm}

\subsection{Interpolation and semi-Lagrangian methods}
As indicated earlier, we use the semi-Lagrangian method to solve equation \eqref{eq:ls} when the velocity field is externally generated, i.e. when it does not depend explicitly on the level-set function itself. Let us rewrite equation \eqref{eq:ls} along the characteristic curve $\underline{\mathbf{X}}(t)$ as:
\be
\left\{
\begin{array}{rcl}
\dfrac{\text{d} \underline{\mathbf{X}}}{\text{d} t} &=& \underline{\mbf{u}}, \\ [3ex]
\dfrac{\text{d} \phi(\underline{\mathbf{X}}(t), t)}{\text{d} t} &=& 0.
\end{array}
\right.
\label{eq:sl}
\ee
The semi-Lagrangian method integrates equations \eqref{eq:sl} backward in time, i.e. starting from the grid $G^{n+1}$ (computed in iteratively as explained later on), we simply write $\phi^{n+1}(\underline{\mathbf{X}}^{n+1}) = \phi(\underline{\mathbf{X}}(t^{n+1}), t^{n+1}) = \phi(\underline{\mathbf{X}}(t^n), t^n) = \phi^n(\underline{\mathbf{X}}_d)$. Here, the characteristic curves are chosen such that $\underline{\mathbf{X}}(t^{n+1})$ are simply the coordinates of grids points of $G^{n+1}$, and $\underline{\mathbf{X}}_d$ are the departure points, which are computed using the second-order midpoint method \cite{Min;Gibou:07:A-second-order-accur}:
\bea
\underline{\mathbf{X}}^\star &=& \underline{\mathbf{X}}^{n+1} - \frac{\Delta t}{2} \underline{\mbf{u}}^{n}(\underline{\mathbf{X}}^n),	   \label{eq:xstar}      \\
\underline{\mathbf{X}}_d     &=& \underline{\mathbf{X}}^{n+1} - \Delta t \underline{\mbf{u}}^{n+\frac{1}{2}}(\underline{\mathbf{X}}^\star), \label{eq:xdeparture}
\eea
where $\underline{\mbf{u}}^{n+\frac{1}{2}}$ is obtained via extrapolation from previous times, i.e.:
\be
\underline{\mbf{u}}^{n+\frac{1}{2}} = \frac{3}{2} \underline{\mbf{u}}^n - \frac{1}{2}\underline{\mbf{u}}^{n-1}. \label{eq:vn_p_half}
\ee
Note that all values at the intermediate point, $\underline{\mathbf{X}}^\star$, and departure point, $\underline{\mathbf{X}}_d$, must be calculated via interpolation from the previous grids $G^{n}$ and $G^{n-1}$. Here, we use the stabilized second-order interpolation for $\phi(\underline{\mathbf{X}}_d)$ and the multi-linear interpolation for $\underline{\mbf{u}}^{n+\frac{1}{2}}(\underline{\mathbf{X}}^\star)$ \cite{Min;Gibou:07:A-second-order-accur}. Although parallelization of the interpolation process on a shared-memory machine is trivial, the same cannot be said for distributed-memory machines. In fact, the parallel interpolation procedure given in Algorithm \ref{alg:interpolation} is probably the most important contribution of this article since the procedure, which is trivial on uniform grids, is challenging in the case of trees because it is not straightforward to identify which processes owns the departure points. Indeed, complications arise because not all departure points will reside in the domain owned by the current process. Moreover, due to the irregular shapes of the partitions, one cannot even ensure they are entirely owned by neighboring processes. At best we can only expect that their locations are bounded by a halo of width $w \le \text{CFL} \: \Delta x_{\min}$ around the local partition, where $x_{\min}$ is the size of the smallest cell in the forest. Naturally, if one enforces $\text{CFL} \le 1$, one can ensure that the halo is bounded by the ghost layer, which significantly simplifies the communication problem. This assumption, however, defeats the purpose of using a semi-Lagrangian approach, whose purpose is to enable large CFL values.

One remedy to this problem, proposed in \cite{Thomas;Cote:95:Massively-parallel-s} for uniform grids, is to increase the size of ghost layer to $\lceil \text{CFL} \rceil$. For large values of the $\text{CFL}$ number, however, this approach can substantially increase the communication volume. Moreover, this simple approach does not work in the process of generating $G^{n+1}$ due to repartitioning. Indeed, $G^{n+1}$ is built iteratively and load balancing is enforced by repartitioning at each sub-iteration. Therefore, after one such sub-iteration, the backtracked points can end up outside of the initial ghost layer. An alternative approach would be to handle local and remote interpolations separately. Our remote interpolation algorithm is composed of three separate phases. In the first phase, which we call buffering, every process searches for all departure points inside the local trees. If the point is owned by a local cell, it is added to a local buffer, otherwise we find the process which owns the point and add the point to a separate buffer belonging to the found rank. Note that searching the point in the local tree is performed recursively using the hierarchal reconstruction (c.f.\ algorithm \ref{alg:reconstruction}). Moreover, the owner's rank is found by computing the Z-index of the point and then using a binary search on the Z-curve. This is already implemented in \texttt{p4est} and explained in details in section 2.5 of \cite{Burstedde;Wilcox;Ghattas:11:p4est:-Scalable-Algo}.

@@

To solve the communication problem, we first compute the adjacency matrix of the communication pattern, i.e. we construct the matrix $A_{P \times P}$, where $P$ is the number of processes, such that
\ben
a_{ij} = 
\left\{
\begin{array}{lr}
1 \hspace{5 mm} \text{if process `\textit{i}' sends a message to process `\textit{j}',} \\
0 \hspace{5 mm} \text{otherwise.}
\end{array}
\right.
\een
Note that this matrix is also distributed among processes, i.e. each row is owned by a separate rank. Next, we compute
\ben
S_i = \sum_j a_{ij} \hspace{5 mm} \text {and} \hspace{5 mm} R_i = \sum_j a_{ji}
\een
where $S_i$ and $R_i$ denote the number of sent and received messages, respectively. While $S_i$ can be computed trivially, a reduction operation is required to compute $R_i$. For instance, this can be achieved using a single \texttt{MPI\_Reduce\_scatter} function call. The last phase of the interpolation procedure involves overlapping the computation of interpolated values for local points with the communication of data between processes. This is done by alternating between local calculations and probing for incoming messages from other processes. The interpolation is finished once the values for all local the points have been calculated and all the remote requests have been processed (see Algorithm \ref{alg:interpolation}).
\begin{algorithm}[htbp]
\caption{$values \gets \texttt{Interpolate (}H, F, \underline{\mathbf{X}}\texttt{)}$: interpolate the value of $F$, defined on the local tree hierarchy $H$, at coordinates $\underline{\mathbf{X}}$}
\begin{algorithmic}[1]
\State $col \gets 0, buff \gets null$ \Comment{Phase I -- buffering}
\For {$\underline{\mathbf{p}} : \underline{\mathbf{X}}$} 
	\State $[r, cell] \gets H.\texttt{search(}\underline{\mathbf{p}}\texttt{)}$ \Comment{search for the owner's rank and cell}
	\If {$r = mpirank$}
		\State $buff[r].\texttt{push\_back(}\underline{\mathbf{p}}, cell\texttt{)}$
	\Else
		\State $buff[r].\texttt{push\_back(}\underline{\mathbf{p}}\texttt{)}$
		\State $col[r] \gets 1$
	\EndIf
\EndFor
\For {$r: mpisize$} \Comment{Phase II -- initiate communication and compute number of messages}
	\If {$col[r]$}
		\State $\texttt{MP\_Isend(}r, buff[r]\texttt{)}$
	\EndIf
\EndFor
\State  $S \gets \texttt{sum(}col\texttt{)}$ 
\State  $R \gets \texttt{MPI\_Reduce\_scatter(}col,\texttt{MPI\_SUM)}$
\State $done \gets false, it \gets buff[mpirank].\texttt{begin()}$
\While {$!done$} \Comment{Phase III -- main loop}
	\If {$it \not= buff[mpirank].\texttt{end()}$}
		\State $values \gets \texttt{process\_local(}it\texttt{)}$ \Comment{process local interpolations}
		\State $\texttt{++}it$
	\EndIf
	\If {$R > 0$} \Comment{process queries sent from remote processes}
		\State $[msg, st] \gets \texttt{MPI\_Iprobe()}$
		\If {$msg$}
			% \State $values \gets \texttt{MPI\_Recv(}st.\texttt{MPI\_SOURCE)}$ \Comment{}
			\State $val\_buff \gets \texttt{process\_queries(}st\texttt{)}$ \Comment{receive, search, and interpolate values}
			\State $\texttt{MPI\_Isend(}st.\texttt{MPI\_SOURCE},val\_buff\texttt{)}$ \Comment{send back interpolated values}
			\State $R\texttt{--}$
		\EndIf
	\EndIf
	\If {$S > 0$} \Comment{process replies sent to our queries}
		\State $[msg, st] \gets \texttt{MPI\_Iprobe()}$
		\If {$msg$}
			\State $values \gets \texttt{process\_replies(}st.\texttt{MPI\_SOURCE)}$ \Comment{receive remotely interpolated values}
			\State $S\texttt{--}$
		\EndIf
	\EndIf
	\State $done \gets S = 0 \: \And \: R = 0 \: \And \: it = buff[mpirank].\texttt{end()}$
\EndWhile
\State \Return $values$
\end{algorithmic}
\label{alg:interpolation}
\end{algorithm}

Using the interpolation Algorithm \ref{alg:interpolation}, we close this section by presenting the final semi-Lagrangian Algorithm \ref{alg:semi-lagrangian}. The basic idea is to start from an initial guess $G^{n+1}_0$ for the grid and modify it using the refinement \eqref{eq:refine} and coarsening \eqref{eq:coarsen} criteria until convergence is obtained. Various options are available for $G^{n+1}_0$. For instance it is possible to start from the macromesh and only perform refinement steps until convergence. This choice, however, is not suitable since the first few iterations do not contain many cells and there is little work for parallelism. Here we simply take the previous grid as the starting point, i.e. $G^{n+1}_0 = G^n$. Note that this iterative process is essentially unavoidable since the grid is based on the values of the level-set function at $t^{n+1}$, which itself is unknown and is to be defined on $G^{n+1}$. Nonetheless the process converges to the final grid in at most $l_{\max}-l_{\min}$ steps where $l_{\min}$ and $l_{\max}$ denote the maximum and minimum depth of all trees in the forest, receptively.
\begin{algorithm}[htbp]
\caption{$[G^{n+1}, \phi^{n+1}] \gets \texttt{SemiLagrangian (}G^n, \phi^n, \underline{\mathbf{u}}^n, \underline{\mathbf{u}}^{n-1}, \text{CFL}\texttt{)}$: update $\phi^{n+1}$ from $\phi^n$ using a semi-Lagrangian scheme and construct the new forest $G^{n+1}$ that is consistent with the zero level-set of $\phi^{n+1}$}
\begin{algorithmic}[1]
\State $\Delta t_l \gets \text{CFL} \times G^n.\texttt{hmin()} / \max \{\underline{\mathbf{u}}^n \} $
\State $\Delta t   \gets \texttt{MPI\_Allreduce(}\Delta t_l,\texttt{MPI\_MIN)}$
\State $H^n \gets \texttt{Reconstruct(} G^n\texttt{)}$ 
\State $G^{n+1}_0 \gets G^n$
\While {$true$}
	\State $\underline{\mathbf{X}}_d  \gets \texttt{ComputeDeparturePoints(}G^{n+1}_0, \underline{\mathbf{u}}^n, \underline{\mathbf{u}}^{n-1}, \Delta t\texttt{)}$ \Comment{using equations \ref{eq:xstar} -- \ref{eq:vn_p_half}}
	\State $\phi^{n+1} \gets \texttt{Interpolate(}H^n, \phi^n, \underline{\textbf{X}}_d\texttt{)}$
	\State $G^{n+1} \gets G^{n+1}_0.\texttt{refine\_and\_coarsen(}\phi^{n+1}\texttt{)}$ \Comment{using equations \ref{eq:refine} and \ref{eq:coarsen} as criteria}
	\If {$G^{n+1} \not= G^{n+1}_0$}
		\State $G^{n+1}.\texttt{partition()}$
		\State $G^{n+1}_0 \gets G^{n+1}$
	\Else
		\State \textbf{break}
	\EndIf
\EndWhile
